---
title: "Sutton-wide Vivacity data check"
author: "Harley Day and John Garnsworthy"
output:
  word_document: default
  pdf_document: default
  html_document:
    df_print: paged
  html_notebook: default
---

I've reproduced the work of John Garnsworthy here to confirm his calculations.

## Import the data from the Excel spreadsheet

Firstly, import the dataset from the Excel spreadsheet. The data we're processing is on the sheet entitled `"all LBS 26 July to 19 Dec"`, and we skip some of the columns which are empty, hence the long `col_types` argument.

```{r}
library(tidyverse)
library(readxl)
library(lubridate)
library(here)
library(ggplot2)
library(tidyr)
library(here)

Sutton_wide_traffic_data <-
  read_excel(
    here("data", "raw_data", "Vivacity data.xlsx"),
    sheet = "all LBS 26 July to 19 Dec",
    col_types = c(
      "date",
      "date",
      "text",
      "text",
      "numeric",
      "numeric",
      "numeric",
      "numeric",
      "numeric",
      "numeric",
      "numeric",
      "numeric",
      "skip",
      "skip",
      "skip",
      "skip",
      "skip"
    ),
    .name_repair = "universal"
  ) %>%
  rename_with(toupper) %>% # Modify all names to uppercase
  rename(DATE = UTC.TIME) %>% # rename the clunky date column to "DATE" for simplicity
  mutate(DAY = wday(DATE, label = TRUE)) %>% # add a "day of the week" column
  write_csv(here("data", "processed_data", "sutton_wide_traffic_data.csv"))
```

This results in a table containing the counts for each mode at each countline for each sensor for each day.

```{r, echo=FALSE}
Sutton_wide_traffic_data
```

Next, we calculate the total number of vehicles of any type in either direction ("in" or "out") each day for all sensors.

```{r}
## define the modes of transport we're interested in analyzing
transport_modes <-
  c("CAR",
    "PEDESTRIAN",
    "CYCLIST",
    "MOTORBIKE",
    "BUS",
    "OGV1",
    "OGV2",
    "LGV")

sum_either_direction_each_day <- Sutton_wide_traffic_data %>%
  group_by(DATE, COUNTLINENAME) %>%
  summarise(across(all_of(transport_modes), sum)) %>%
  group_by(DATE, COUNTLINENAME) %>%
  mutate(TOTAL = sum(c_across(cols = all_of(transport_modes))))

sum_either_direction_each_day
```

## Filtering out days with zero events

John spotted that some sensors record no traffic of any mode at all on some days, from which we can reasonably infer the sensor was not active. We calculate the number of presumed inactive days during the date range as follows. Some of the sensors recorded traffic on every single day in our date range, so they do not appear in this table.
```{r}
number_days_with_zero_observations <- sum_either_direction_each_day %>%
  filter(TOTAL==0) %>%
  group_by(COUNTLINENAME) %>%
  summarise(n = n())
number_days_with_zero_observations
```

We filter out the inactive days as follows.

```{r}
## filter out those days with zero observations as they're probably days when the sensor was inactive
sum_nonzero_either_direction_each_day <- sum_either_direction_each_day %>%
  filter(TOTAL>0)
```

## Johns's summary

Finally, we calculate the totals and averages for each mode at each location. The numbers I calculte here match John's exactly. I conclude John's spreadsheet is free of errors.

```{r}
## filter summary statistics by date
summary_stats_26Jul2021_to_19Dec2021 <- sum_nonzero_either_direction_each_day %>%
  group_by(COUNTLINENAME) %>%
  filter(DATE <= as_date("2021-12-29") && DATE >= as_date("2021-07-26")) %>%
  summarise(across(all_of(transport_modes), list(total = sum, mean = mean)))

summary_stats_26Jul2021_to_19Dec2021
```

## Remaining issues

Some of the sensors seem to be measuring suspiciously low values, even if the total number of events is greater than zero. For example, even after we're removed the days with no events (see filtering out inactive days above), sensor 130 "L1_VickersCl_road_s130" records the following total counts per day for each mode.

```{r}
sum_nonzero_either_direction_each_day %>%
  filter(COUNTLINENAME=="L1_VickersCl_road_s130") %>%
  pivot_longer(all_of(transport_modes), names_to = "MODE", values_to = "COUNT") %>%
  ggplot(aes(
    x = DATE,
    y = COUNT,
    group = MODE,
    colour = MODE
  )) + geom_line() +
  labs(title = "L1_VickersCl_road_s130")
```
This is not the only sensor with a suspiciousely low reading for a period of time. Many of the sensors that appear on the table of those with "inactive days" have this same issue. Here's "L12_LindberghRd_road_and_LHS_path_s140".

```{r}
sum_nonzero_either_direction_each_day %>%
filter(COUNTLINENAME=="L12_LindberghRd_road_and_LHS_path_s140") %>%
pivot_longer(all_of(transport_modes), names_to = "MODE", values_to = "COUNT") %>%
ggplot(aes(
  x = DATE,
  y = COUNT,
  group = MODE,
  colour = MODE
)) + geom_line() +
labs(title = "L12_LindberghRd_road_and_LHS_path_s140")
```
  
These low reading periods are not being filtered out by John's method, so I suggest we ask Vivacity what's causing this stange behaviour, and visually check each of the sensor traces to work out which periods of time we can trust. The "L1_VickersCl_road_s130" sensor, for example, seems to perk up in early September, giving the kind of levels for each mode we might expext. That said, there are still large spikes in these data, which I suspect are some sort of artifact we also need to filter out.